On Thue Numbers

### 2016-07-30 draft.tex
final!

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage[normalem]{ulem} 
\usepackage{wrapfig}
\usepackage{caption, subcaption}
\usepackage{todonotes}
\usepackage[style=ieee, backend=bibtex, natbib=true]{biblatex}
\usepackage{standalone}
\author{Theresa Eimer}
\title{On Thue Numbers}
\date{}

@@

\bibliography{library.bib}
\begin{document}
\renewcommand{\figurename}{Fig.}
\maketitle
\thispagestyle{empty}
\newpage
\tableofcontents
\thispagestyle{empty}
\newpage

\section{Introduction}
In 1906, Axel Thue showed the construction of an infinite word where no two adjacent parts are be the same using only three letters \citep{Thue1906}. This concept of nonrepetitiveness has since been applied to other fields, probably most successfully in graph theory. The nonrepetitiveness of graph colorings and the number of colors needed to construct nonrepetitive colorings have been a topic since Alon et. al. introduced the idea in 2002 \citep{Alon2002}. This thesis aims to give an overview over some results of the results, both in vertex- and edge-colorings. 
\newline
First, some general preliminaries in graph and complexity theory will be covered. 
\newline
Then, in the third section, the formal definition of nonrepetitiveness for graph colorings and the complexity of the corresponding decision problem will be shown. 
\newline
The fourth section is about edge-colorings, more specifically Manin's proof that the Thue Number problem is $\Sigma^p_2$-complete \citep{Manin2008}. This proof will be examined in detail as it's probably the most important result for nonrepetitive edge-colorings as well as generally applicable, which isn't the case for most results specific to vertex-colorings.
\newline
Section five, therefore, is intended to be a bit of an overview over some relevant results when it comes to nonrepetitive vertex-colorings, though it certainly doesn't aim to be complete. The first part introduces two randomized algorithms, one for finding repetitive paths of a specific length and the other constructing a nonrepetitive coloring. After that, bounds for two classes of graphs that have received particular attention will be presented: planar graphs and trees. 
\newline
Finally, section six will mention other applications and variants of nonrepetitiveness and the Thue Number.
\newpage
\section{Preliminaries}
\subsection{Graph Theory}

\begin{definition}
A graph $G$ is defined as two finite sets V and E, representing the vertices and edges of $G$ respectively, so that $E$ is a subset of $ V ^ 2$, meaning that each edge from $E$ connects two vertices from $V$ \citep[p. 1]{Bollobas1998}.
\end{definition} 
This is denoted as $G = (V, E)$. Hereafter, any graph $G$ will refer to an undirected and unweighted graph unless mentioned otherwise. Thus, edges can be traversed in both directions. Furthermore, all graphs will be simple graphs, not containing multi-edges (more than one edge from a vertex $v_i$ to another vertex $v_j$) or loops (edges starting from and ending at the same vertex) \citep[pp. 3-4]{Gross2013}. 
\begin{definition}
A graph $G' = (V', E')$ is called a subgraph of $G$ if $V'$ is a subset of $V$ and $E'$ is a subset of $E$ \citep[p. 2]{Bollobas1998}.
\end{definition}
\begin{definition}
Two vertices $v_i$ and $v_j$ are \textit{adjacent}, if there is an edge $e_k$ that connects the two. Two edges are adjacent if they share one vertex \citep[p. 1-2]{Bollobas1998}. 
\end{definition}
\begin{definition}
A path on a graph is a sequence of the form 
\begin{center}
	$ W = v_0 e_0 v_1 e_1...e_{n-1} v_n $
\end{center}
where, starting from an initial vertex $v_0$, vertices and edges alternate until the final vertex $v_n$ is reached.
\end{definition}
In a simple graph, a path can also be represented by either just its vertices \citep[p. 10]{Gross2013} or edges \citep[p. 4]{Bollobas1998}. The length $l$ of a path is the number of edges it contains \citep[p. 5]{Bollobas1998}.
\begin{definition}
An open path is a path that visits no vertex more than once.
\end{definition} Thus cycles on the path are not possible (see 'trail' in \citep[p. 10]{Gross2013}). Open paths are sometimes called walks. 
\begin{definition}An open path through every vertex is called a Hamiltonian path \citep[p. 14]{Bollobas1998}.
\end{definition}
\begin{definition}
An edge-coloring of a graph $G$ is a function that matches each edge to a color contained in $C$, the set of colors.  A vertex-coloring is a function matching the vertices instead of the edges to the colors in $C$ \citep[pp. 7-8]{Gross2013}.
\end{definition} 
\newpage
\begin{wrapfigure}{r}{4cm}
	\includestandalone{pics/clique}
	\caption{8-Clique}
\end{wrapfigure}
\par There are some special graph structures that will become important later on: the \textit{Clique} and the \textit{Hypercube}. 
\begin{definition}
A Clique is a subgraph of a graph G, in which every two vertices $v_i$, $v_j$ are adjacent (see figure 1).  A Clique containing k vertices is denoted as $k-Clique$ \citep[pp. 21, 111]{Gross2013}.
\end{definition}
\begin{definition}
A Hypercube is an $n$-dimensional Graph (denoted n-Hypercube) that can be constructed by labeling a set of vertices of size $2^n$ with a different element of $\{ 0, 1 \}^n$ each and connecting the vertices that differ in one position (see figure 2) \citep[p. 60]{Bollobas1998}.
\end{definition}
If $n > 1$, the construction of the $(n+1)$-Hypercube consist essentially of two $n$-Hypercubes being connected (see figure 3). Both for the Hypercube and Clique, the number of edges connected to a vertex is the same for all vertices, e.g. 7 for a 8-Clique or 4 for a 4-Hypercube.
\newline

\par
\begin{figure}[h]
\begin{minipage}{.3\textwidth}
	\includestandalone{pics/hypercube_small}
	\captionof{figure}{3-Hypercube}
\end{minipage}
\begin{minipage}{.5\textwidth}
	\includestandalone{pics/hypercube_big}
	\centering
	\captionsetup[figure]{oneside,margin={2cm,0cm}}
	\captionof{figure}{4-Hypercube constructed out of two 3-Hypercubes}

@@

An alternative way to define $NP$ is that $NP$ contains all problems for which the a solution can be verified in polynomial time. \citep[p. 59]{Goldreich} Its counterpart, the complexity class of $coNP$, contains all problems where an instance that is not a solution can be verified as such in polynomial time \citep[p. 154-155]{Goldreich}. 

@@

The further levels are recursively defined using \textit{oracles}. If a Turing Machine can use an oracle, it means that in a single computational step, a $L$ oracle will decide if a string on the Turing Machine's tape is contained in $L$. With that in mind, the definitions go as follows:

@@

	$\Sigma^p_{i+1} = \{$ $L$ | $L$ is accepted by a Non-deterministic Turing Machine with a $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$

@@

	$\Pi^p_{i+1} = \{$ $L$ | the complement of $L$ is accepted by a Non-deterministic Turing Machine with a $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$

@@

	$\Delta^p_{i+1} = \{$ $L$ | $L$ is accepted by a Deterministic Turing Machine with a $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$

@@

Three problems that will be reduced from later on in this thesis are $SAT$, $3SAT$ and $\forall \exists 3SAT$. The later two are special cases of $SAT$, the boolean satisfiability problem. 

@@

$SAT$ is the language containing all boolean formulas in conjunctive normal form that are satisfiable. 

@@

$SAT$ is in $NP$ and in fact $NP-hard$, meaning that every problem in $NP$ can be reduced to $SAT$. Both of those attributes combined make it $NP-complete$. \citep[p. 338-346]{Davis1983} 

@@

$3SAT$ is also $NP-complete$. \citep[p. 347]{Davis1983} 

@@

It's $\Pi^p_2$-complete \citep{Schaefer2002}.

\newpage
\section{Nonrepetitive graph colorings}
\subsection{Nonrepetitiveness}
Before talking about Thue number type problems, it makes sense to look at non-repetitiveness first, as it is the defining concept behind the  Thue number. Nonrepetitive here means square-free. This can be applied to different fields like graph theory, geometry or number theory \citep{Grytczuk2008}, but it was first introduced in the process of constructing a square-free word by Axel Thue in 1906. 
\begin{definition} 
A word $w$ is square-free if there is no part $x$ of the word that occurs two times in a row.
\end{definition} 
Let the words $a$ and $b$ each consist of three different substrings, $v$, $x$ and $y$, then $a = vxxy$ is not square-free, but $b = vxy$ is \citep{Thue1906}.
\newline
In graph theory, nonrepetitiveness is relevant with respect to nonrepetitive vertex- \citep{Marx2009} and edge-colorings of graphs. 
\begin{definition} 
A coloring $e$ of a graph $G$ is nonrepetitive, if there is no open path whose vertex or edge colors contain a color sequence $x$ that is repeated two or more times in a row.
\end{definition} 
To put it differently: a coloring of a graph is nonrepetitive if the coloring of every possible open path on the graph is nonrepetitive \citep{Alon2002}. To give a baseline for the complexity of further problems concerning nonrepetitive graph colorings, first the complexity of determining whether a vertex- or edge coloring is nonrepetitive will be shown.

@@

Deciding whether a vertex-coloring of a graph is nonrepetitive is coNP-complete. \citep{Marx2009a}

@@

The problem of deciding whether a vertex-coloring of a graph $G(V, E)$ is nonrepetitive has been proven to be $coNP-hard$ by reducing from the Hamiltonian path problem, which is $NP-complete$. The idea behind Marx's and Schäfer's proof is to construct a graph H from G with two parts, the second one having a path for every possible path in G that is as long a Hamiltonian path would be ($n$ vertices with $n = |E|$), and the first part representing the vertices of $G$ with a subgraph each. $H$ is colored in such a way that a repetitively colored path would have to go through both parts with the coloring from the first path repeating itself in the second path. If a path through the second half of $H$, however, corresponds to a path in $G$ that is not a Hamiltonian path but a path that visits at least one vertex more than once, the subgraph for that vertex in the first half of $H$ would have to also be visited more than once - but that would mean that the path through the first half is not an open path. Thus $H$ is colored nonrepetitively only if $G$ doesn't have a Hamiltonian path. This implies that the problem of deciding if a vertex-coloring is nonrepetitive is $coNP-hard$.

@@

@@

The construction of $H$ goes as follows: for the first part, for every vertex $i \leq n$ ($n = |E|$) of $G$, a $K_{2, n}$ graph (fig. 4) is added. It consists of two vertices, colored $a$ and $b$ respectively, with $n$ vertices between them, each connected to both the $a$- and $b$-vertex. As one vertex is added for each vertex in $G$, the vertices are colored $c_{i,j}$, $i$ being the vertex in $G$ for which the subgraph was added and $j$ ranging from 1 to $n$. The $[b]$ vertices are now connected to $[a]$ vertex, $z$, colored $c$, that marks the transition between the two parts of $H$. Lastly, to connect the different subgraphs, for every $[b]$ vertex, $d_{i,j}$-colored vertices are added. A $d_{i, j}$ vertex is connected to the $[b]$ vertex of the subgraph with the same value for $i$ and all $[a]$ vertices for $j$ between 1 and $n$ where $j \neq i$.  So now all $[a]$ and $[b]$ vertices are connected to all other $K_{2,n}$ subgraphs of the first part of $H$.

@@

For the second part, for every combination of $i$ and $j$ ($1 \leq i, j \leq n$), a path containing three vertices colored a, $c_{i,j}$ and $b$ respectively are added to $H$. Where $j = 1$, the $[a]$ vertex is connected to the middle vertex $z$. Between the $[b]$ vertices for $j = 1$ and the $[a]$ vertices for $j = 2$, $d_{i,j}$-colored vertices are added. $i$ here is the same as in three vertex path to which the new vertex becomes connected via the $[b]$ vertex, $j$ the number of every vertex the $i$-th vertex in $G$ is connected to by an edge in $E$. Those $[d]$ vertices are then connected to the $[a]$ of the three vertices path corresponding to the $j$ value in their color. This way, the connections in $G$ are mimicked in the structure of $H$. This is repeated for every $j$ until all the three vertex subgraphs have been connected. The last step is to connect the $[b]$ vertices of the paths where $j = n$ to a $[c]$ end vertex $z'$. 

@@

@@

Now $H$ has a first part that allows any path through the graph to be traced through the coloring, a vertex of a different color separating the two halves, and a second part whose connections are dependent on the edges of $G$. In the first and second path, $a$ and $b$ are the only non-unique colors and vertices colored $a$ or $b$ are never adjacent, so a repetitive path has run through both parts, that means it needs to pass the $c$-colored vertex $z$, and because of this also needs to reach the end vertex as it is the only other vertex colored with $c$. Thus, the only possible square path would look like this: $w = QzQ'z'$ with $Q$ and $Q'$ and of course $z$ and $z'$ having the same coloring. To get from vertex $z$ to vertex $z'$, $w$ has to pass through $n$ vertices colored $c_{i,j}$ for some $i,j$, visiting the same colors in the first part of $H$. As all vertices colored $c_{i,j}$ with the same value for $i$ are contained in the same subgraph of the first half of $H$, $w$ can only visit one $c_{i,j}$ colored vertex for each value of $i$. This is only possible if $G$ has a Hamiltonian path, because then, by construction of the second part, there is a path through the second half of $H$ that doesn't visit any two vertices with the same value for $i$. The coloring of this path can simply be copied for a path through the first part. If $G$ doesn't have a Hamiltonian path, there is no path 
through the second part of $H$ without having to visit two different vertices with the same value for $i$, but to get the same coloring for the first part, the path there would have to pass through one of the $K_{2,n}$ subgraphs twice, therefore it is not an open path. $F$ is an example of a graph without a Hamiltonian path, so it will not be possible to find a path of the form $w = QzQ'z'$, the example below from Marx's and Schäfer's paper, however, does contain a square path. The graph it's based on is $F'$. 
\todo{Add pic}
	\caption{F'}

@@

The corresponding problem for edge-colorings is in the same complexity class and the proof is somewhat similar, they are however not directly related and one can't be inferred from the other \citep{Marx2009a}. The reason Manin's proof might seem familiar is that it's also based on the idea of a path with two parts that are colored identically to form the only possible repetitive colored path. The problem he reduces from is $3SAT$, corresponding to which a graph $G$ is constructed. $G$ has a so called snout, a sequence of vertices and edges with the sole purpose of providing the colors for the second part to mimic. The second part of $G$ then consist of a variable gadget for each variable that splits into a positive and negative path variant and a clause gadget for each clause. The clause gadgets are connected to the variable gadgets in the same way the variables occur in the clause; if a variable is negated, the clause gadget is connected to a vertex in the negative path of the variable gadget, if it isn't, the clause gadget is connected to the positive path. As a result, the only way to pass through the second half of G with an open path, is find an assignment satisfying the formula, traversing the variable gadgets opposite to that assignment and then the clause gadgets corresponding to it. If there is no satisfying assignment, at least one of the edges from the clause gadgets will need to pass through a vertex in the variable gadget that has already been visited. Thus a coloring in $G$ is only repetitive if the $3SAT$ problem it's constructed from is satisfiable - and nonrepetitive if it isn't.

@@

Again, the construction of $G$ only takes a polynomial number of steps, therefore the reduction proofs that the problem of deciding the nonrepetitiveness of an edge-coloring is $coNP-hard$. 

@@

The reasoning here is the same as in section 3.2., thus the problem is $coNP-complete$.

@@

\section{Thue Numbers for graphs}
\subsection{The Thue number and Thue chromatic Number}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{hyphenat}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage[normalem]{ulem} 
\usepackage{wrapfig}
\usepackage{caption, subcaption}
\usepackage{pdfpages}
\usepackage[style=ieee, backend=bibtex, natbib=true]{biblatex}
\usepackage{standalone}
\author{Theresa Eimer}
\title{On Thue Numbers}
\date{}

\newcommand*{\xdash}[1][3em]{\rule[0.5ex]{#1}{0.55pt}}

@@

\bibliography{library.bib}
\begin{document}
\renewcommand{\figurename}{Fig.}
\includepdf[pages=-]{deckblatt.pdf}
\thispagestyle{empty}
\newpage
\tableofcontents
\thispagestyle{empty}
\newpage

\section{Introduction}
In 1906, Axel Thue showed the construction of an infinite word where no two adjacent parts are the same using only three letters \citep{Thue1906}. This concept of nonrepetitiveness has since been applied to other fields, probably most interestingly in graph theory. The nonrepetitiveness of graph colorings and the number of colors needed to construct such nonrepetitive colorings have been a topic since Alon et. al. introduced the idea in \citep{Alon2002}. They focused on coloring the vertices of graphs in ways to that no partial coloring of a path repeats itself immediately, similar to Thue's word. It's also possible to delve into the nonrepetitiveness of colorings of the edges instead of the vertices of graphs. This thesis aims to give an overview of some of the results, both in vertex- and edge-colorings with the main task being the analysis of Fedor Manin's proof of complexity for the Thue Number problem. The Thue Number $\pi(G)$ of a graph $G$ is the minimum of colors that is needed for a nonrepetitive edge-coloring of $G$. Manin showed this problem to be $\Sigma^p_2-complete$, but even though his proof has been referenced in \citep{Marx2009} and \citep{Schaefer2002}, it hasn't been published. Most of it's correct, although there are some minor mistakes and many places in the proof where the motivation or reason for a particular construction is missing, even going as far as leaving out parts of the construction process. This thesis attempted to fix both the incorrect parts of the proof and add to the construction in such a way that it might become more accessible and easier to understand.
\newline
First, though, some general preliminaries in graph and complexity theory will be covered to give the reader a background in these fields.
\newline
Then, in the third section, the formal definition of nonrepetitiveness for graph colorings and the complexity of the corresponding decision problems will be given. These problems serve both as an introduction to the Thue Number proof, in fact one of them is its predecessor, and to nonrepetitive colorings in general. Furthermore, comparing the two might make clear why the Thue Number problem for edge-colorings has been solved, but the corresponding vertex-coloring problem hasn't yet.
\newline
The fourth section is about edge-colorings, more specifically Manin's proof that the Thue Number problem is $\Sigma^p_2$-complete \citep{Manin2008}. This proof will be examined in detail as it's probably the most important result for nonrepetitive edge-colorings as well as generally applicable, which isn't the case for most results specific to vertex-colorings. In this proof, as well as Manin's proof in section three, some inconsistencies and missing links have been resolved and the proof has also been fleshed out a good bit more than in the original paper.
\newpage
Section five, then, is intended to give a bit of an overview of some relevant results when it comes to nonrepetitive vertex-colorings, though it certainly doesn't aim to be complete. The first part introduces two randomized algorithms, one for finding repetitive paths of a specific length and the other constructing a nonrepetitive coloring. After that, bounds for two classes of graphs that have received particular attention will be presented: planar graphs and trees. The results here, with maybe the exception of trees, won't be presented in depth as a lot of them have been found using probabilistic methods which go beyond the scope of this thesis. They're mentioned to show how diverse methods and results for different colorings and graphs are, not as an exhaustive explanation of nonrepetitive vertex-colorings.
\newline
Finally, section six will mention other applications and variants of nonrepetitiveness and Thue Number type problems in other mathematical fields.
\newpage

\section{Preliminaries}

Before going into the specifics of nonrepetitive colorings, this section provides the background knowledge necessary for the two main fields of this thesis: graph theory and complexity. Basic as well as more specific terminology that will become important later will be defined here. Broader and more detailed explanation can be found in \citep{Bollobas1998}, \citep{Gross2013} and \citep{Davis1983} that were used for this section. Note that this section is focused on graph and complexity theory, but in places some general mathematical background might still be necessary to understand parts of this thesis.

\subsection{Graph Theory}

\begin{definition}
A graph $G$ is defined as two finite sets V and E, representing the vertices and edges of $G$ respectively, so that $E$ is a subset of $ V \times V$, meaning that each edge from $E$ connects two vertices from $V$ \citep[p. 1]{Bollobas1998}.
\end{definition} 
This is denoted as $G = (V, E)$. Hereafter, any graph $G$ will refer to an undirected and unweighted graph unless mentioned otherwise. Thus, edges can be traversed in both directions. Furthermore, all graphs will be simple graphs, not containing multi-edges (more than one edge from a vertex $v_i$ to another vertex $v_j$) or loops (edges starting from and ending at the same vertex) \citep[pp. 3-4]{Gross2013}. 
\begin{definition}
A graph $G' = (V', E')$ is called a subgraph of $G$ if $V'$ is a subset of $V$ and $E'$ is a subset of $E$ \citep[p. 2]{Bollobas1998}.
\end{definition}
\begin{definition}
The maximal degree $\Delta$ of a graph is the maximal number of edges connected to a single vertex \citep[p. 4]{Bollobas1998}.
\end{definition}
There is also a minimum degree, but it won't be relevant for the topics covered here, so degree will always refer to the maximum degree.
\begin{definition}
Two vertices $v_i$ and $v_j$ are \textit{adjacent}, if there is an edge $e_k$ that connects the two. Two edges are adjacent if they share one vertex \citep[p. 1-2]{Bollobas1998}. 
\end{definition}
\begin{definition}
A path on a graph is a sequence of the form 
\begin{center}
	$ W = v_0 e_0 v_1 e_1...e_{n-1} v_n $
\end{center}
where, starting from an initial vertex $v_0$, vertices and edges alternate until the final vertex $v_n$ is reached.
\end{definition}
In a simple graph, a path can also be represented by either just its vertices \citep[p. 10]{Gross2013} or edges \citep[p. 4]{Bollobas1998}. The length $l$ of a path is the number of edges it contains \citep[p. 5]{Bollobas1998}.
\newpage
\begin{definition}
An open path is a path that visits no vertex more than once.
\end{definition} Thus cycles on the path are not possible (see 'trail' in \citep[p. 10]{Gross2013}). Open paths are sometimes called walks. 
\begin{definition}An open path through every vertex is called a Hamiltonian path \citep[p. 14]{Bollobas1998}.
\end{definition}
This shouldn't be confused with the Hamiltonian circle, a Hamiltonian path visits every vertex exactly once whereas a Hamiltonian circle then goes on to visit the first vertex in the path again.
\begin{definition}
An edge-coloring of a graph $G$ is a function that matches each edge to a color contained in $C$, the set of colors.  A vertex-coloring is a function matching the vertices instead of the edges to the colors in $C$ \citep[pp. 7-8]{Gross2013}. 
\end{definition} 
\begin{wrapfigure}{r}{4cm}
	\includestandalone{pics/clique}
	\caption{8-Clique}
\end{wrapfigure}
The color of an edge $e$ in a coloring $C$ will be denoted as $C(e)$. An edge of color $c$ whose other attributes aren't relevant will be referred to as $[c]$.
\newline

\par There are some special graph structures that will become important later on: the \textit{Clique} and the \textit{Hypercube}. 
\begin{definition}
A Clique is a graph, in which every two vertices $v_i$, $v_j$ are adjacent (see figure 1).  A Clique containing k vertices is denoted as $k-Clique$ \citep[pp. 21, 111]{Gross2013}.
\end{definition}

As all vertices are adjacent in an $n$-Clique, there are $n-1$ edges per vertex. The degree of all vertices is therefore the same in this graph. An example of a Clique can be seen in fig. 1.
\newpage
\begin{definition}
A Hypercube is an $n$-dimensional Graph (denoted n-Hypercube) that can be constructed by labeling a set of vertices of size $2^n$ with a different element of $\{ 0, 1 \}^n$ each and connecting the vertices that differ in one position (see fig. 2) \citep[p. 60]{Bollobas1998}.
\end{definition}
If $n > 1$, the construction of the $(n+1)$-Hypercube consist essentially of two $n$-Hypercubes being connected (see fig. 3). Both for the Hypercube and Clique, the number of edges connected to a vertex is the same for all vertices, e.g. 7 for a 8-Clique or 4 for a 4-Hypercube.
\newline

\par
\begin{figure}[h]
\begin{minipage}{.4\textwidth}
	\includestandalone{pics/hypercube_small}
	\captionof{figure}{3-Hypercube}
\end{minipage}
\begin{minipage}{.5\textwidth}
	\includestandalone{pics/hypercube_big}
	\centering
	\captionsetup[figure]{oneside,margin={2cm,0cm}}
	\captionof{figure}{4-Hypercube constructed out of two 3-Hypercubes}

@@

An alternative way to define $NP$ is that $NP$ contains all problems for which a solution can be verified in polynomial time \citep[p. 59]{Goldreich}. Its counterpart, the complexity class of $coNP$, contains all problems where an instance that is not a solution can be verified as such in polynomial time \citep[p. 154-155]{Goldreich}. 

@@

The further levels are recursively defined using \textit{oracles}. If a Turing Machine can use an oracle, it means that in a single computational step, an oracle for a language $L$ will decide if a string on the Turing Machine's tape is in $L$. With that in mind, the definitions go as follows:

@@

	$\Sigma^p_{i+1} = \{$ $L$ | $L$ is accepted by a Non-deterministic Turing Machine with an $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$

@@

	$\Pi^p_{i+1} = \{$ $L$ | the complement of $L$ is accepted by a Non-deterministic Turing Machine with an $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$

@@

	$\Delta^p_{i+1} = \{$ $L$ | $L$ is accepted by a Deterministic Turing Machine with an $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$

@@

Three problems that will be reduced from later on in this thesis are $SAT$, $3SAT$ and $\forall \exists 3SAT$. The later two are special cases of $SAT$, the Boolean Satisfiability Problem. 

@@

$SAT$ is the language containing all satisfiable boolean formulas in conjunctive normal form. 

@@

$SAT$ is in $NP$ and in fact $NP-hard$, meaning that every problem in $NP$ can be reduced to $SAT$. Both of those attributes combined make it $NP-complete$ \citep[p. 338-346]{Davis1983}.

@@

$3SAT$ is also $NP-complete$ \citep[p. 347]{Davis1983}.

@@

$\forall \exists 3SAT$ is $\Pi^p_2$-complete \citep{Schaefer2002}.

\newpage
\section{Nonrepetitive graph colorings}
This section will introduce the notion of nonrepetitiveness before delving into problems stemming from it. It includes two smaller proofs, both showing the complexity of deciding whether a coloring is nonrepetitive, one for vertex- and one for edge-colorings. The vertex-coloring proof is taken from \citep{Marx2009} and not much, beside an additional example and showing why it's in $coNP$, has been added. There are hardly any proofs of complexity when it comes to vertex-colorings, so this one, as the most basic problem for Thue Number type problems, gives a good point of orientation of how more complicated proofs might look like. Furthermore, the second proof \citep{Manin2008} was directly extended to proof the complexity of the Thue Number problem, so comparing the two also gives an idea of why the vertex-coloring version of Thue Number can't be proven as easily with the same idea. The second proof itself, of course, leads up to the bigger Thue Number proof, but has a few inconsistencies that have been remedied here. Its concept and graph construction will be the basis for the next section.

\subsection{Definition of Nonrepetitiveness}
Before talking about Thue number type problems, it makes sense to look at non-repetitiveness first, as it is the defining concept behind the  Thue number. Nonrepetitive here means square-free. This can be applied to different fields like graph theory, geometry or number theory \citep{Grytczuk2008}, but it was first introduced in the process of constructing a square-free word by Axel Thue in 1906. 
\begin{definition} 
A word $w$ is square-free if there is no part $x$ of the word that occurs two times in a row.
\end{definition} 
Let the words $a$ and $b$ each consist of three different substrings, $v$, $x$ and $y$, then $a = vxxy$ is not square-free, but $b = vxy$ is \citep{Thue1906}.
\newline
In graph theory, nonrepetitiveness is relevant with respect to nonrepetitive vertex- \citep{Marx2009} and edge-colorings of graphs. 
\begin{definition} 
A coloring $e$ of a graph $G$ is nonrepetitive if there is no open path whose vertex or edge colors contain a color sequence $x$ that is repeated two or more times in a row.
\end{definition} 
To put it differently: a coloring of a graph is nonrepetitive if the coloring of every possible open path on the graph is nonrepetitive \citep{Alon2002}. To give a baseline for the complexity of further problems concerning nonrepetitive graph colorings, first the complexity of determining whether a vertex- or edge coloring is nonrepetitive will be shown.

@@

Deciding whether a vertex-coloring of a graph is nonrepetitive is coNP-complete \citep{Marx2009}.

@@

The problem of deciding whether a vertex-coloring of a graph $G(V, E)$ is nonrepetitive has been proven to be $coNP-hard$ by reducing from the Hamiltonian path problem, which is $NP-complete$. The idea behind Marx's and Schäfer's proof is to construct a graph $H$ from $G$ with two parts, the first part representing the vertices of $G$ with a subgraph each and the second one having a path for every possible path in G that is as long as a Hamiltonian path would be ($n$ vertices with $n = |E|$). $H$ is colored in such a way that a repetitively colored path would have to go through both parts with the coloring from the first path repeating itself in the second path. If a path through the second half of $H$, however, corresponds to a path in $G$ that is not a Hamiltonian path but a path that visits at least one vertex more than once, the subgraph for that vertex in the first half of $H$ would have to also be visited more than once - meaning that the path through the first half is not an open path. Thus $H$ is colored nonrepetitively only if $G$ doesn't have a Hamiltonian path. This implies that the problem of deciding whether a vertex-coloring is nonrepetitive is $coNP-hard$.

@@

\par

@@

The construction of $H$ goes as follows: for the first part, for every vertex $i \leq n$ ($n = |E|$) of $G$, a $K_{2, n}$ graph (fig. 4) is added. It consists of two vertices, colored $a$ and $b$ respectively, with $n$ vertices between them, each connected to both the $a$- and $b$-vertex. As one vertex is added for each vertex in $G$, the vertices are colored $c_{i,j}$, $i$ being the vertex in $G$ for which the subgraph was added and $j$ ranging from 1 to $n$. The $[b]$ vertices are now connected to $[a]$ vertex, $z$, colored $c$, that marks the transition between the two parts of $H$. Lastly, to connect the different subgraphs, for every $[b]$ vertex, $d_{i,j}$-colored vertices are added. A $d_{i, j}$ vertex is connected to the $[b]$ vertex of the subgraph with the same value for $i$ and to all $[a]$ vertices for $j$ between 1 and $n$ where $j \neq i$.  So now all $[a]$ and $[b]$ vertices are connected to all other $K_{2,n}$ subgraphs of the first part of $H$. An example of this construction for a simple graph (fig. 5) is shown in fig. 6.

@@

For the second part, for every combination of $i$ and $j$ ($1 \leq i, j \leq n$), a path containing three vertices colored a, $c_{i,j}$ and $b$ in this order is added to $H$. Where $j = 1$, the $[a]$ vertex is connected to the middle vertex $z$. Between the $[b]$ vertices for $j = 1$ and the $[a]$ vertices for $j = 2$, $d_{i,j}$-colored vertices are added. The value for $i$ is the same as in the three vertex path to which the new vertex becomes connected via the $[b]$ vertex and $j$ takes on the values of every vertex the $i$-th vertex in $G$ is connected to by an edge in $E$. Those $[d]$ vertices are then connected to the $[a]$ of the three vertices path corresponding to the $j$ value in their color. This way, the connections in $G$ are mimicked in the structure of $H$. This is repeated for every $j$ until all the three vertex subgraphs have been connected. The last step is to connect the $[b]$ vertices of the paths where $j = n$ to a $[c]$ end vertex $z'$. Fig. 7 shows the completed example construction.

@@

Now $H$ has a first part that allows any path through the graph to be mimicked by the coloring, a vertex of a different color separating the two parts of $H$, and a second part whose connections are dependent on the edges of $G$. In the first and second part respectively, $a$ and $b$ are the only non-unique colors and vertices colored $a$ or $b$ are never adjacent. Thus a repetitive path has to run through both parts, that means it needs to pass the $c$-colored vertex $z$. Because of this, it also needs to reach the end vertex, as it is the only other vertex colored with $c$. So the only possible square path would look like this: $w = QzQ'z'$ with $Q$ and $Q'$ and of course $z$ and $z'$ having the same coloring. To get from vertex $z$ to vertex $z'$, $w$ has to pass through $n$ vertices colored $c_{i,j}$ for some $i,j$, visiting the same colors in the first part of $H$. As all vertices colored $c_{i,j}$ with the same value for $i$ are contained in the same subgraph of the first half of $H$, $w$ can only visit one $c_{i,j}$ colored vertex for each value of $i$. This is only possible if $G$ has a Hamiltonian path, because then, by construction of the second part, there is a path through the second half of $H$ that doesn't visit any two vertices with the same value for $i$. The coloring of this path can simply be copied for a path through the first part. If $G$ doesn't have a Hamiltonian path, there is no path 
through the second part of $H$ without having to visit two different vertices with the same value for $i$. To get the same coloring for the first part, however, the path there would have to pass through one of the $K_{2,n}$ subgraphs twice. Therefore it is not an open path. 
\begin{minipage}[c]{0.3\linewidth}
\begin{minipage}[c]{0.8\linewidth}
	\includegraphics[scale=0.3]{pics/ex_marx.png}
\begin{minipage}[t]{0.2\linewidth}
	\caption{$F'$}
\begin{minipage}[t]{0.8\linewidth}
	\caption{Example construction for $F'$ \citep{Marx2009}}

@@

$F$ is an example of a graph without a Hamiltonian path, so it will not be possible to find a path of the form $w = QzQ'z'$. The example given in fig. 9 (based on the graph from fig. 8) from Marx's and Schäfer's paper, though, has a square path.

@@

The corresponding problem for edge-colorings is in the same complexity class and the proof is somewhat similar. They are, however, not directly related and one can't be inferred from the other \citep{Marx2009}. The reason Manin's proof might seem familiar is that it's also based on the idea of a path with two parts that are colored identically to form the only possible repetitive colored path. The problem he reduces from is $3SAT$, corresponding to which a graph $G$ is constructed. $G$ has a so called snout, a sequence of vertices and edges with the sole purpose of providing the colors for the second part to mimic. The second part of $G$ then consist of a variable gadget for each variable that splits into a positive and negative path variant and a clause gadget for each clause. The clause gadgets are connected to the variable gadgets in the same way the variables occur in the clause; if a variable is negated, the clause gadget is connected to a vertex in the negative path of the variable gadget, if it isn't, the clause gadget is connected to the positive path. As a result, the only way to pass through the second half of G with an open path, is to find an assignment satisfying the formula, traversing the variable gadgets opposite to that assignment and then the clause gadgets corresponding to it. If there is no satisfying assignment, at least one of the edges from the clause gadgets will need to pass through a vertex in the variable gadget that has already been visited. Thus a coloring in $G$ is only repetitive if the $3SAT$ problem it's constructed from is satisfiable - and nonrepetitive if it isn't.

@@

Again, the construction of $G$ only takes a polynomial number of steps, therefore the reduction proves that the problem of deciding the nonrepetitiveness of an edge-coloring is $coNP-hard$. 

@@

The reasoning here is the same as in section 3.2., thus making the problem $coNP-complete$.

@@

\section{The complexity of the  Thue Number problem}
